---
DGE analysis
---

```{r setup}
suppressMessages({
  library(tidyverse)
  library(DESeq2)
  library(AnnotationHub)
  library(tximeta)
  library(clusterProfiler)
})

volcano_path <- '../plot/volcanos/'
go_path <- '../plot/Gene_ontology/'
venn_path <- '/media/alessio/Data/hypoxia/RNAseq2025/plot/venn/'

Volcanoplot <- function(df, title) {
  
  up <- nrow(df[df$DEG == 'UP', 'DEG'])
  down <- nrow(df[df$DEG == 'DOWN', 'DEG'])
  
  df %>% 
    ggplot(aes(x = log2FoldChange, y = -log10(pvalue), color = DEG)) +
    geom_point(alpha = 0.7) +
    scale_color_manual(values = c("UP" = "red", "DOWN" = "blue", "ns" = "black")) +
    labs(title = title,
         subtitle = paste0('UP: ',up,
                           '\nDOWN: ',down),
         color = NULL) +
    theme_bw()
}

L2FC = 0.75
```

# load data and classify UP and DOWN regulated
```{r}
files <- list.files('../DEGs_tabs', full.names = T)

DEGs_results <- tibble(files = files)
DEGs_results <- DEGs_results %>% 
  dplyr::mutate(deseq_res = map(files, \(x) read_delim(x))) %>% 
  dplyr::mutate(files = str_remove(basename(files), '.tsv')) %>% 
  dplyr::mutate(deseq_res = map(
    deseq_res, \(x) x %>% 
      dplyr::mutate(DEG = case_when(
        (padj < 0.05 & log2FoldChange > L2FC) ~ 'UP',
        (padj < 0.05 & log2FoldChange < -L2FC) ~ 'DOWN',
        TRUE ~ 'ns'
      )
      )
  )
  )
```

# Check of the difference of the two clones with their controls at 0h 
### venn of shared genes, volcano and pathway enrichment
```{r}
H4KO_vs_ctrl_0h <- DEGs_results %>% 
  dplyr::filter(files %in% c('H4KO_125_vs_ctrl_0h',
                             'H4KO_26_vs_ctrl_0h')
                )

# volcano and enrichment dotplot sep clones
genes_up_H4KO_vs_ctrl <- list()

for (i in H4KO_vs_ctrl_0h$files) {
  df <- H4KO_vs_ctrl_0h %>% 
    dplyr::filter(files == i) %>% 
    unnest()
  
  # volcano
  P <- Volcanoplot(df, i) + lims(x = c(-15, 15))
  print(P)
  
  genes_up <- na.omit(unique(df[df$DEG == 'UP', ]$symbol))
  
  genes_up_H4KO_vs_ctrl[[i]] <- genes_up
  # GO enrichment
  GO_enrichment <- enrichGO(genes_up,
                            ont = "BP",
                            keyType = "SYMBOL",
                            OrgDb = "org.Hs.eg.db")

  dt <- dotplot(GO_enrichment, title = i)
  print(dt)
  # save plots
  ggsave(P, filename = paste0(volcano_path,i,'_volcano.png'), height = 6, width = 6)
  ggsave(dt, filename = paste0(go_path,i,'_go.png'), height = 6, width = 6)
}

VennDiagram::venn.diagram(genes_up_H4KO_vs_ctrl, 
                          disable.logging = T, 
                          filename = paste0(venn_path,'HDAC4_KO_vs_ctrl_clones_sep_intersect.png'),
                          margin = 0.1, imagetype = 'png')

write_lines(Reduce(intersect, genes_up_H4KO_vs_ctrl), '../HDAC4_KO_vs_ctrl_normoxia_intersect_clones_sep.txt')
```

# Check hypoxia response of the clones separated
```{r}
# prepare data
Vs_normoxia <- DEGs_results %>% 
  dplyr::filter(str_detect(files, 'interaction|ctrl', negate = T)) %>% 
  dplyr::mutate(clone = str_remove(files, '_0h|_6h|_24h|_48h')) %>% 
  dplyr::mutate(time = case_when(
    str_detect(files, '6h') ~ '6h',
    str_detect(files, '24h') ~ '24h',
    TRUE ~ '48h'
  )) %>% 
  arrange(time)

# Barplot with summary of number of DEGs genes
Vs_normoxia_summary_DEGs_number <- Vs_normoxia %>% 
  unnest() %>% 
  dplyr::select(clone, DEG, time) %>%
  group_by(clone, DEG, time) %>% 
  summarize(count = n()) %>% 
  arrange(time) %>% 
  dplyr::filter(DEG != 'ns')

#Plot
barplot_degs <- Vs_normoxia_summary_DEGs_number %>% 
  dplyr::mutate(count = ifelse(DEG == "UP", count, -count)) %>%
  dplyr::mutate(time = fct_relevel(time, c("6h", "24h", "48h"))
  ) %>%
  ggplot(aes(x = time, y = count, fill = DEG)) +
  geom_col(color = "black", width = 0.5) +
  scale_fill_manual(values = c("UP" = "#eb1719", "DOWN" = "#4baae8")) +
  geom_text(aes(x = time,
                y = ifelse(DEG == "UP", count + 200, count -200),
                label = ifelse(DEG == "UP", count, -count))) +
  facet_wrap(~ clone, scales = "free_x", ncol = 4) +
  theme_bw() +
  labs(x = NULL,
       y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y=element_blank())

barplot_degs
```
### save barplot
```{r}
ggsave(barplot_degs, filename = '../plot/barplot_degs_hypoxia.png', height = 6, width = 9)
```


### volcano and Gene Ontology enrichemnt
```{r}
GO_enrichment_list <- list()


for (i in Vs_normoxia$files) {
  df <- Vs_normoxia %>% 
    dplyr::filter(files == i) %>% 
    unnest()
  
  P <- Volcanoplot(df, i) + lims(x = c(-15, 15))
  print(P)
  
  genes_up <- df[df$DEG == 'UP', ]$symbol
  
  GO_enrichment <- enrichGO(genes_up, 
                            ont = "BP",
                            keyType = "SYMBOL",
                            OrgDb = "org.Hs.eg.db")
  
  GO_enrichment_list[[i]] <- GO_enrichment  
  
  dt <- dotplot(GO_enrichment, title = i)
  
  print(dt)
  ggsave(P, filename = paste0(volcano_path,i,'_volcano.png'), height = 6, width = 6)
  ggsave(dt, filename = paste0(go_path,i,'_go.png'), height = 6, width = 6)
  
}
```

# summarize the results of hypoxia time response gene ontology
```{r}
GO_enrichment_list_long <- GO_enrichment_list %>% 
  enframe(name = 'sample', value = 'GO') %>% 
  dplyr::mutate(GO = map(GO, \(x) as.data.frame(x))) %>% 
  unnest() %>% 
  dplyr::filter(str_detect(Description, 'Hypox|hypox|xygen')) %>% 
  mutate(
    sample = fct_relevel(
      sample, 
      c(
        "Cas9_125_6h", "Cas9_125_24h", "Cas9_125_48h",
        "Cas9_26_6h", "Cas9_26_24h", "Cas9_26_48h",
        "H4KO_125_6h", "H4KO_125_24h", "H4KO_125_48h",
        "H4KO_26_6h", "H4KO_26_24h", "H4KO_26_48h"
      )
    )
  )


plot_summary_hypoxia_go <- GO_enrichment_list_long %>% 
  ggplot(aes(x = sample, y = Description)) +
  geom_point(aes(size = Count, color = -log10(qvalue))) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ggsave(plot_summary_hypoxia_go, filename = '../plot/bubbleplot_summary_hypoxia_go.png', height = 5, width = 8)
```

# interaction terms analysis
### we want to find genes commonly de-regulated in both HDAC4 KO clones
### we do this on both the deseq results of interaction taken the clones together or separated and then taking the common at the end
```{r}
L2FC = 0.5

# clones 125 and 26 together treated as replicates
pathx <- '/media/alessio/Data/hypoxia/RNAseq2025/interaction_hypoxia_repclones'
hypoxia_interaction <- DEGs_results %>% 
  dplyr::filter(str_detect(files, '^H4KO_vs_ctrl(.*)interaction$')) %>% 
  dplyr::mutate(time = case_when(
    str_detect(files, '6h') ~ '6h',
    str_detect(files, '24h') ~ '24h',
    str_detect(files, '48h') ~ '48h'
  ))  %>% 
  dplyr::mutate(deseq_res = map(
    deseq_res, \(x) x %>% 
      dplyr::mutate(DEG = case_when(
        (padj < 0.05 & log2FoldChange > L2FC) ~ 'UP',
        (padj < 0.05 & log2FoldChange < -L2FC) ~ 'DOWN',
        TRUE ~ 'ns'
      )
      )
  )
  )

UP_genes <- list()
for (i in unique(hypoxia_interaction$time)) {
  df <- hypoxia_interaction %>%
    dplyr::filter(time == i)

  print(paste0('Time is: ',i))

  UP <- df %>%
      unnest(deseq_res) %>%
      dplyr::filter(DEG == 'UP')
  
  print(paste0('number of Upreg for ',i,' is: ',nrow(UP)))
  
  UP_genes[[i]] <- na.omit(UP$symbol)
  
}

DOWN_genes <- list()
for (i in unique(hypoxia_interaction$time)) {
  df <- hypoxia_interaction %>%
    dplyr::filter(time == i)

  print(paste0('Time is: ',i))

  DOWN <- df %>%
      unnest(deseq_res) %>%
      dplyr::filter(DEG == 'DOWN')
  
  print(paste0('number of Upreg for ',i,' is: ',nrow(DOWN)))
  
  DOWN_genes[[i]] <- na.omit(DOWN$symbol)
}

# save to txt 
for (i in names(UP_genes)) {
  file = paste0(pathx,'/','up_interaction_',i,'.txt')
  write_lines(UP_genes[[i]], file = file)
}

for (i in names(DOWN_genes)) {
  file = paste0(pathx,'/','down_interaction_',i,'.txt')
  write_lines(DOWN_genes[[i]], file = file)
}

# clones separated
pathx <- '/media/alessio/Data/hypoxia/RNAseq2025/interaction_hypoxia_common'

hypoxia_interaction_sep_clones <- DEGs_results %>% 
  dplyr::filter(str_detect(files, '^H4KO_(26|125)(.*)interaction$')) %>% 
  dplyr::mutate(time = case_when(
    str_detect(files, '6h') ~ '6h',
    str_detect(files, '24h') ~ '24h',
    str_detect(files, '48h') ~ '48h'
  ))  %>% 
  dplyr::mutate(deseq_res = map(
    deseq_res, \(x) x %>% 
      dplyr::mutate(DEG = case_when(
        (padj < 0.05 & log2FoldChange > L2FC) ~ 'UP',
        (padj < 0.05 & log2FoldChange < -L2FC) ~ 'DOWN',
        TRUE ~ 'ns'
      )
      )
  )
  )


# Upregulated common of interaction
Up_intersect <- list()
for (i in unique(hypoxia_interaction_sep_clones$time)) {
  df <- hypoxia_interaction_sep_clones %>%
    dplyr::filter(time == i)

  print(paste0('Time is: ',i))

  common_UP <- list()
  for (j in unique(df$files)) {
    UP <- df %>%
      dplyr::filter(files == j) %>%
      unnest(deseq_res) %>%
      dplyr::filter(DEG == 'UP')

    print(paste0('number of Upreg for ',j,' is: ',nrow(UP)))
    common_UP[[j]] <- UP$symbol
  }
  common_UP <- na.omit(Reduce(intersect, common_UP))

  print(paste0('Commonly upregulated are: ',length(common_UP)))

  Up_intersect[[i]] <- common_UP
}

# Down regulated common of interaction
Down_intersect <- list()
for (i in unique(hypoxia_interaction_sep_clones$time)) {
  df <- hypoxia_interaction_sep_clones %>%
    dplyr::filter(time == i)

  print(paste0('Time is: ',i))

  common_DOWN <- list()
  for (j in unique(df$files)) {
    DOWN <- df %>%
      dplyr::filter(files == j) %>%
      unnest(deseq_res) %>%
      dplyr::filter(DEG == 'DOWN')

    print(paste0('number of Downreg for ',j,' is: ',nrow(DOWN)))
    common_DOWN[[j]] <- DOWN$symbol
  }
  common_DOWN <- na.omit(Reduce(intersect, common_DOWN))

  print(paste0('Commonly Downregulated are: ',length(common_DOWN)))

  Down_intersect[[i]] <- common_DOWN
}

Up_intersect
Down_intersect

for (i in names(Up_intersect)) {
  file = paste0(pathx,'/','up_common_interaction_',i,'.txt')
  write_lines(Up_intersect[[i]], file = file)
}

for (i in names(Down_intersect)) {
  file = paste0(pathx,'/','down_common_interaction_',i,'.txt')
  write_lines(Down_intersect[[i]], file = file)
}
```
what happens here is that we have some genes that have different behavior in HDAC4 KO than in KO,
For example the Up-regulated does not mean that they are Up regulated in absolute, but they have an increased de-regulation
in HDAC4 KO than in Cas9, this can be both UP and DOWN. 







